plyranges: A grammar of genomic data transformation

### 2018-09-20 paper.Rmd

To support merging, our algebra recasts finding overlaps or 
nearest neighbors between two
genomic regions as variants of the relational join operator. A join
acts on two GRanges objects: _x_ and _y_. The join operator
is relational in the sense that metadata from the _x_ and _y_
ranges are retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figures 
\ref{fig:olaps-fig} and \ref{fig:nn-fig}). We can further restrict the
matching by whether _y_ is completely _within_ the _x_, and ensure that 
matching ranges have the same direction (strand) using the _directed_ suffix.

```{r olaps-fig, echo = FALSE, out.width="\\textwidth", fig.cap="Illustration of the three overlap join operators. Each join takes query and subject range as input (black and light gray rectangles, respectively). An index for the join is computed, returning a Hits object, which contains the indices of where the subject overlaps the query range. This index is used to expand the query ranges by where it was 'hit' by the subject ranges. The join semantics alter what is returned: for an \\textbf{inner} join the query range is returned for each match, for an \\textbf{intersect} the intersection is taken between overlapping ranges, and for a \\textbf{left} join all query ranges are returned even if the subject range does not overlap them. This principle is generally applied through the `plyranges` DSL for both overlaps and nearest neighbour operations."}
knitr::include_graphics("diagrams/diagrams-002.png")
```

>>>>

To support merging, our algebra recasts finding overlaps or 
nearest neighbors between two
genomic regions as variants of the relational join operator. A join
acts on two GRanges objects: _x_ and _y_. The join operator
is relational in the sense that metadata from the _x_ and _y_
ranges are retained in the joined range.  All join operators in the
`plyranges` DSL generate a set of hits based on overlap or proximity
of ranges and use those hits to merge the two datasets in different
ways. There are four supported matching algorithms: _overlap_,
_nearest_, _precede_, and _follow_ (figure 
\ref{fig:olaps-fig}). We can further restrict the
matching by whether the query is completely _within_ the subject, and
adding the _directed_ suffix ensures that matching ranges have the
same direction (strand).

```{r olaps-fig, echo = FALSE, out.width="\\textwidth", fig.cap="Illustration of the three overlap join operators. Each join takes two GRanges objects, \\textit{x} and \\textit{y} as input. A 'Hits' object for the join is computed which consists of two components. The first component contains the indices of the ranges in \\textit{x} that have been overlapped (the rectangles of \\textit{x} that cross the orange lines). The second component consists of the indices of the ranges in \\textit{y} that overlap the ranges in \\textit{x}. In this case a range in \\textit{y} overlaps the ranges in \\textit{x} three times, so the index is repeated three times. Te resulting 'Hits' object is used to modify \\textit{x} by where it was 'hit' by \\textit{y}. This principle of is applied generally in the \\texttt{plyranges} DSL for both overlap and nearest neighbour operations. The join semantics alter what is returned: \\textbf{A}: for an \\textbf{inner} join the \textit{x} ranges that are overlapped by \\textit{y} are returned. \\textbf{B}  For an \\textbf{intersect} join the intersection is taken between the overlapped \\textit{x} ranges and the \\textit{y} ranges. \\textbf{C} For the left join \\textbf{left} join all \\textit{x} ranges are returned regardless of whether they are overlapped by \\textit{y}."}
knitr::include_graphics("diagrams/olap-fig.png")
```
